### 2022-02-14 ack.tex
Typos

First and foremost, I would like to thank my advisor, Martin Odersky, for his help and support throughout my PhD study, and for providing me with this unique opportunity to work on my favorite language, Scala.

@@

I would like to thank all my colleagues from LAMP, LARA and Scala Center.
You were the source of so many interesting discussions, cheerful moments, and great memories, it was really a pleasure to work along your side.
Aggelos Biboudis,
Aleksander Boruch-Gruszecki,
Allan Renucci,
Anatolii Kmetiuk,
Darja Jovanovic,
Denys Shabalin,
Dmitry Petrashko,
Fabien Salvi,
Felix Mulder,
Fengyun Liu,
Guillaume Martres,
Guillaume Masse,
Jad Hamza,
Jamie Thompson,
Jonathan Brachthäuser,
Jorge Vicente,
Julien Richard-Foy,
Manohar Jonnalagedda,
Matt Bovel,
Natascha Fontana,
Nicolas Alexander Stucki,
Ólafur Páll Geirsson,
Paolo Giarrusso,
Romain Edelmann,
Sandro Stucki,
Sébastien Doeraene,
Simon Guilloud, and
Travis Lee---thank you!
If it wasn't for you, I would have not have gotten here.

A special thanks to Georg Schmid for being such an amazing friend, colleague and co-author over the past five years.
I will never forget what we went through together, the paper deadlines, the hacking sessions, the bike rides, all games...
It's crazy how much we did and how much we learned together.
And of course, thank you Mia Primorac for being there for all the fun parts!

I would like to thank my parents, Soledad and Christian, and my little sister, Alicia, for being there for me all these years.
It was truly a pleasure to have you close by throughout my PhD.
I must also thank the latest arrival in family, Leila, for being such an endless source of joy and cuteness.

Last but not the least, I would like to thank Dragana Milovancevic for being always there on my side, for being so kind and patient, and for her tremendous help in putting the ideas of this thesis into text.
You are the best.

>>>>

I would like to thank my advisor, Martin Odersky, for his help and support throughout my PhD studies, and for providing me with this unique opportunity to work on my favorite language.

@@

I would like to thank all my colleagues from LAMP, LARA and Scala Center.
You were the source of so many interesting discussions, cheerful moments, and great memories, it was really a pleasure to work along your side.
Aggelos Biboudis,
Aleksander Boruch-Gruszecki,
Allan Renucci,
Anatolii Kmetiuk,
Darja Jovanovic,
Denys Shabalin,
Dmitry Petrashko,
Dragana Milovancevic,
Fabien Salvi,
Felix Mulder,
Fengyun Liu,
Georg Schmid,
Guillaume Martres,
Guillaume Masse,
Jad Hamza,
Jamie Thompson,
Jonathan Brachthäuser,
Jorge Vicente,
Julien Richard-Foy,
Manohar Jonnalagedda,
Matthieu Bovel,
Maxime Kjaer,
Natascha Fontana,
Nicolas Stucki,
Ólafur Geirsson,
Paolo Giarrusso,
Romain Edelmann,
Sandro Stucki,
Sébastien Doeraene,
Simon Guilloud, and
Travis Lee---thank you!
If it wasn't for you, I would have not have gotten here.

A special thanks to Georg Schmid for being such an amazing friend, colleague and co-author over the past five years.
I will never forget what we went through together, the paper deadlines, the bike rides, the hacking sessions, the games...
It's crazy how much we did and learned together.
And of course, thank you Mia Primorac for being there for all the fun parts!

I would like to thank my parents, Soledad and Christian, and my little sister, Alicia, for being there for me all these years.
It was truly a chance and a pleasure to have you close by throughout my PhD.
I must also thank the latest arrival in family, Leila, for being such an endless source of joy, cuteness and distraction.

Last but not the least, I would like to thank Dragana for being always there on my side, for being so kind and patient, and for her tremendous help in putting the ideas of this thesis into text.
You are the best.

### 2022-02-14 article.tex
Type-Safe Regular Expressions: related work
Typos

\subsubsection*{Preliminary Lemmas}
Our proof begins with preliminary technical lemmas: \prelemma{Permutation}, \prelemma{Weakening}, \prelemma{Strengthening}, \prelemma{Substitution}.
We omit stating these lemmas as they are entirely standard yet relatively lengthy given that they span the three relations of our system: typing, subtyping, and disjointness.
As usual, their proofs follow by mutual inductions on derivations.

@@

\section{Related work}
\label{sec:related-work}

>>>>

\subsubsection*{Preliminary Lemmas}
Our proof begins with preliminary technical lemmas:
%
\input{proofs/permutation.tex}
\input{proofs/weakening.tex}
\input{proofs/strengthening.tex}
\input{proofs/substitution.tex}
%
These lemmas are entirely standard, and as usual, their proofs follow by mutual inductions on derivations.

@@

\section{Related Work}
\label{sec:related-work}

### 2022-02-14 symposium.tex
Typos

The implementation of the |Regex| class is straightforward.
Similar to its standard library counterpart, that class uses an instance of |java.util.regex.Pattern| to match the given input against the given regular expression, and to extract the content of the capturing groups when the matching succeeds.
These operations operations take place within the |unapply| method, which is the way to define custom pattern matching extractors in Scala \citep{emir2007matching}.

@@

\noindent
Here, we instantiates |Regex|'s type parameter to |P = (String, Option[String])|, which we then use in |unapply|'s return type to obtain a precise representation of the regex's capturing groups.
The resulting code is type-safe: if we change the pattern to omit the option unpacking (replacing |Some(f)| by |f|), the compiler would raise a type error on the call to |.size| (size is defined on strings but not options).

In the following section, we explain our technique to analysis regular expressions and statically compute a type-level representation of the regex's capturing groups (|P| in the example above).

@@

\noindent
The result of that type-level computation takes the form of a Scala tuple with |String| and |Option[String]| elements, depending of the nullability analysis.

@@

Be reassured, this is not an oversight!
Regular expressions with trailing backslashes are invalid and result in runtime crashes (reported as "unexpected internal error").
Even though detecting these types of errors at compile time is out of the scope of our library, doing so is still desirable.
If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting that error at runtime.

@@

Our first implementation of |Compile| treats every opening parenthesis as the start of a capturing group, which does not honor the syntax of Java's regular expression.
Indeed, Java also supports several other special constructs which start with an opening parenthesis, for instance:

@@

We handle nested groups by updated our algorithm to operate in two different modes while iterating through the regex's characters.
Our algorithm can either be outside of an optional group (|Opt=0|), in which case it computes nullability of newly encountered groups via |IsNullable|, or inside an optional group (|Opt>0|), in which case it treats every group as nullable.

@@

The second version of our runtime component takes a completely different approach.
Instead of analyzing regular expressions at runtime, to reconstitute type-level analysis results, we use implicit resolution to \emph{synthesize} a type-specialized runtime.
That synthesis is type-directed: it generates a single-purpose program that sanitized the output of the regex engine by following the shape of the tuple type computed at the type level.

Our implementation consists of one type class, |Sanitizer|, and three implicit definitions, which will guide the compiler into generating the correct |Sanitizer[T]|, for any tuple |T| with |String| and |Option[String]| elements (here, |T| is the results of our type-level analysis).
The |Sanitizer| type class defines a single method that mutates an array in place to warp nullable elements in options:

@@

To assess the correctness of our library, we run our implementation against the QT3TS test suite \citep{w3c1994xquery}, which consists of about 1700 test cases for regular expressions \citep[perl-tests.xml]{w3c1994qt3ts}.
We select the subset of those tests that are positive (they match successfully) and exercises capturing groups, totaling 183 test cases, which we then collated to form our benchmark suite.
We use these benchmarks to evaluate the performance of our library in terms of compilation times and execution times.

@@

We obtained these results by averaging the measurements obtained over 2 hour runs, for each data point, which we executed on an i7-7700K Processor running Oracle JVM 1.8.0\_212 and Linux.
With these settings, we obtains margins of errors below \textpm1\%, with 99.9\% confidence (omitted in \Cref{fig:regexBars}).
We run our benchmarks on the latest version of Scala~3 at the time of writing (3.1.2).

\paragraph{Compilation Times} Our library in its code-duplicated variant adds a total of 53~ms over the baseline (an average of +0.3~ms per regex).
This corresponds to the cost of match type reduction.
The implicit-based variant adds another 21~ms (+0.1~ms per regex), which corresponds to the cost implicit resolution.
Both variants are thus relatively cheap, which validates their utility for practical uses.
When it comes to compilation times, it would be misleading to compare increments relative to the overall time, given that the compiler spends a large portion of its time outside of type checking \citep[2.11.3]{petrashko2017design}.

\paragraph{Execution Times} At runtime, our code-duplicated variant adds 51~\textmu s over the baseline.
Although this difference is three orders of magnitudes smaller than at compile-time, it corresponds to a 25\% increase, which is a rather substantial price to pay for redoing an already-performed analysis.
The implicit-based variant gets rid of this cost by synthesizing a type-specialized runtime (at the cost of increased compilation time).
To our surprise, our implicit-based runtime outperforms Scala's standard library implementation.
We suspect that this difference is due to the poor interaction between the JVM's just-in-time compiler and the code generated for variadic extractors (|unapplySeq|, used by the Std implementation).
Our implicit-based implementation appears to be more prone to inlining and partial evaluation since it does not involve generic sequences, which could explain the slight gain in performance.

@@

The Haskell ecosystem contains several packages for regular expressions, with various levels of static safety.
The regex-applicative package stands on the safe side of that spectrum: it allows programmers to write regular expressions using a parser combinator library, which entirely removes the need to reference capturing groups \citep{cheplyaka2011regex}.
The main drawback of this solution lies in the library's learning curves: programmers already familiar with POSIX regular expressions~\citep{ieee2018the} are constrained to learn a new syntax, which is arguably more complicated.

Stephanie Weirich presented, on several occasions \citep{weirich2014examples}, a regular expression library as an example of a dependently-typed program written in Haskell.
The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
To make that possible, her implementation makes heavy use of Haskell's advanced type system features and requires no less than 20 language extensions to compile.
In contrast, our implementation does not use metaprogramming and compiles with vanilla Scala~3.
An important conceptual difference is that our approach only focuses on type checking (we rely on Java's regular expressions at runtime), whereas her library also contains a complete regular expression engine.

@@

\paragraph{Future Work}
We propose to extend our approach with a dedicated data type for the regex alternative operator.
In our current design, we map every optional regex operator to an |Option[String]|, which does not always accurately reflect the structure of regular expressions.
For instance, in \lstinline!"(A)(B)|(C)"! (which matches either |AB| or |C|), our system represents capturing groups as a 3-tuple, |(Option[String],| |Option[String],| |Option[String])|. This representation fails to account for the fact that the first and second groups have identical nullability status, and that this status is opposite to the nullability of the third group.
Instead, we could represent those capturing groups as |Either[(String,| |String),| |String]|, which is isomorphic to the regular expression.

>>>>

The implementation of the |Regex| class is straightforward.
Similar to its standard library counterpart, that class uses an instance of |java.util.regex.Pattern| to match the given input against the given regular expression, and to extract the content of the capturing groups when the matching succeeds.
These operations take place within the |unapply| method, which is the way to define custom pattern matching extractors in Scala \citep{emir2007matching}.

@@

\noindent
Here, we instantiate |Regex|'s type parameter to |P = (String, Option[String])|, which we then use in |unapply|'s return type to obtain a precise representation of the regex's capturing groups.
The resulting code is type-safe: if we change the pattern to omit the option unpacking (replacing |Some(f)| by |f|), the compiler would raise a type error on the call to |.size| (size is defined on strings but not options).

In the following section, we explain our technique to analyze regular expressions and statically compute a type-level representation of the regex's capturing groups (|P| in the example above).

@@

\noindent
The result of that type-level computation takes the form of a Scala tuple with |String| and |Option[String]| elements, depending on the nullability analysis.

@@

Be reassured, this is not an oversight!
Regular expressions with trailing backslashes are invalid and result in runtime crashes (reported as "unexpected internal error").
Even though detecting these types of errors at compile time is out of the scope of our library, doing so is still desirable.
If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting than error at run-time.

@@

Our first implementation of |Compile| treats every opening parenthesis as the start of a capturing group, which does not honor the syntax of Java's regular expression.
Indeed, Java also supports several other special constructs that start with an opening parenthesis, for instance:

@@

We handle nested groups by updating our algorithm to operate in two different modes while iterating through the regex's characters.
Our algorithm can either be outside of an optional group (|Opt=0|), in which case it computes nullability of newly encountered groups via |IsNullable|, or inside an optional group (|Opt>0|), in which case it treats every group as nullable.

@@

The second version of our runtime component takes a completely different approach.
Instead of analyzing regular expressions at run-time, to reconstitute type-level analysis results, we use implicit resolution to \emph{synthesize} a type-specialized runtime.
That synthesis is type-directed: it generates a single-purpose program that sanitizes the output of the regex engine by following the shape of the tuple type computed at the type level.

Our implementation consists of one type class, |Sanitizer|, and three implicit definitions, which will guide the compiler into generating the correct |Sanitizer[T]|, for any tuple |T| with |String| and |Option[String]| elements (here, |T| is the result of our type-level analysis).
The |Sanitizer| type class defines a single method that mutates an array in place to put nullable elements in options:

@@

To assess the correctness of our library, we run our implementation against the QT3TS test suite \citep{w3c1994xquery}, which consists of about 1700 test cases for regular expressions \citep[perl-tests.xml]{w3c1994qt3ts}.
We select the subset of those tests that are positive (they match successfully) and exercise capturing groups, totaling 183 test cases, which we then collated to form our benchmark suite.
We use these benchmarks to evaluate the performance of our library in terms of compilation times and execution times.

@@

We obtained these results by averaging the measurements obtained over 2-hour runs, for each data point, which we executed on an i7-7700K Processor running Oracle JVM 1.8.0\_212 and Linux.
With these settings, we obtain margins of errors below \textpm1\%, with 99.9\% confidence (omitted in \Cref{fig:regexBars}).
We run our benchmarks on the latest version of Scala~3 at the time of writing (3.1.2).

\paragraph{Compilation Times} Our library in its code-duplicated variant adds a total of 53~ms over the baseline (an average of +0.3~ms per regex).
This corresponds to the cost of match type reduction.
The implicit-based variant adds another 21~ms (+0.1~ms per regex), which corresponds to the cost of implicit resolution.
Both variants are thus relatively cheap, which validates their utility for practical uses.
When it comes to compilation times, it would be misleading to compare increments relative to the overall time, given that the compiler spends a large portion of its time outside of type checking \citep[2.11.3]{petrashko2017design}.

\paragraph{Execution Times} At runtime, our code-duplicated variant adds 51~\textmu s over the baseline.
Although this difference is three orders of magnitude smaller than at compile-time, it corresponds to a 25\% increase, which is a rather substantial price to pay for redoing an already-performed analysis.
The implicit-based variant gets rid of this cost by synthesizing a type-specialized runtime (at the cost of increased compilation time).
To our surprise, our implicit-based runtime outperforms Scala's standard library implementation.
We suspect that this difference is due to the poor interaction between the JVM's just-in-time compiler and the code generated for variadic extractors (|unapplySeq|, used by the Std implementation).
Our implicit-based implementation appears to be more prone to inlining and partial evaluation since it does not involve generic sequences, which could explain the slight gain in performance.

@@

The Haskell ecosystem contains several packages for regular expressions, with various levels of static safety.
The regex-applicative package stands on the safe side of that spectrum: it allows programmers to write regular expressions using a parser combinator library, which entirely removes the need to reference capturing groups \citep{cheplyaka2011regex}.
The main drawback of this solution lies in the library's learning curve: programmers already familiar with POSIX regular expressions~\citep{ieee2018the} are forced to learn a new syntax, which is arguably more complicated.

Stephanie Weirich presented, on several occasions \citep{weirich2014examples}, a regular expression library as an example of a dependently-typed program written in Haskell.
The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
To make that possible, her implementation makes heavy use of Haskell's advanced type system features and requires no less than 20 language extensions to compile.
In contrast, our implementation does not use metaprogramming and compiles with vanilla Scala~3.
An important conceptual difference is that our approach only focuses on type checking (we rely on Java's regular expressions at run-time), whereas her library also contains a complete regular expression engine.

@@

\paragraph{Future Work}
We propose to extend our approach with a dedicated data type for the regex alternative operator.
In our current design, we map every capturing group to an |Option[String]|, which does not always accurately reflect the structure of regular expressions.
For instance, in \lstinline!"(A)(B)|(C)"! (which matches either |AB| or |C|), our system represents the capturing groups as a 3-tuple, |(Option[String],| |Option[String],| |Option[String])|. This representation fails to account for the fact that the first two groups have identical nullability status, and that this status is opposite to the nullability of the third group.
Instead, we could represent those capturing groups as |Either[(String,| |String),| |String]|, which is isomorphic to the structure of the regular expression.

### 2022-02-14 thesis.tex
Typos

In this chapter, we evaluate the performance of the various type-level programming techniques introduced in prior chapters.
Specifically, we are interested in the scalability of each technique when confronted with large type-level programs, in terms of compilation times (\Cref{sec:compilation-time}) and binary sizes (\Cref{sec:binary-size}).
In \Cref{sec:the-timing-of-match-type-reductions}, we discuss how the timing of match type reduction impacts compilation times.

@@

We evaluate the performance of the 3 type-level programming techniques presented in this dissertation by comparing them on a set of 4 benchmarks.
Each benchmark consists of a simple type-level operation, which we implemented once using implicits (\Cref{chap:ab-using-implicits}), once using generalized singletons (\Cref{chap:generalizing-singleton-types}), and once using match types (\Cref{chap:match-types}).

@@

  \item \textbf{Remove}\quad
  Removes the last element of a heterogeneous list (\Cref{sec:hlists-remove}).

  \item \textbf{Join}\quad
  Computes the resulting schema after performing a self-join along the last column of a database table (\Cref{sec:motivating-example}).

  \item \textbf{Reduce}\quad
  Computes the dimension of a multidimensional array after dimensional reduction along all axes (\Cref{subsubsec:reduction}).
\end{enumerate}

We obtained the measurements presented in this chapter by averaging 60 independent compilations on a warm compiler, which we executed on an i7-7700K Processor running Oracle JVM~1.8.0\_212 and Linux.
For the implicit and match type benchmarks, we used the latest version of Scala~3 at the time of writing (3.1.2).
For the generalized singleton benchmarks, we used our prototype implementation, which lives in a branch of the dotty-staging repository\footnote{\texttt{git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7}}.

@@

\Cref{fig:timingGraphsZoomedOut} compares the compilation time of implicits, generalized singletons and match types on our benchmark suite.
The compilation time of implicit-based benchmarks dominates the graphs on each figures, and suggests a quadratic complexity.
These results imply that implicits are ill-suited for large scale type-level computations: spending tens of seconds on a single operation is clearly out of the question for practical applications.

@@

The same measurements are shown in \Cref{fig:timingGraphsZoomedIn}, but with millisecond y-axes, to focus on the comparison of non-implicit curves.
Each graph follows a similar trend, with singleton-based implementations outperforming their match-type-based counterparts by about a factor of two.
We speculate that this difference is due to the overhead of subtyping, which in inherent to the design of match type.
Our prototype for generalized singletons uses a dedicated type evaluation procedure that is specialized for the task and thus faster.

We obtained the data points of \Cref{fig:timingGraphsZoomedOut} and~\ref{fig:timingGraphsZoomedIn} using Java Microbenchmark Harness (JMH) with 60 seconds of warmup, 60 seconds of measurement and a single JVM fork.
As a result, the resulting measurements are precise (99.9\% confidence intervals vary between $\pm$ 0.5 and $\pm$ 2 milliseconds), but subject to variations caused by the unpredictability of the JVM's just in time compiler, which explains the bumps visible in \Cref{fig:timingGraphsZoomedIn}.

@@

In \Cref{fig:bytecodeSizeTable}, we show the amount of additional JVM Bytecode generated per benchmark size increment.
The binary sizes of singleton- and match-type-based implementations are stable with respect to benchmark input size, which is expected given that our benchmarks consist of type-level-only operations.

The binary sizes of implicit-based implementations grow linearly with the input size, by 6 bytes per input size increment for smaller benchmarks (Concat and Remove), and by 12 and 18 bytes for the larger benchmarks (Reduce and Join, respectively).
To put those numbers in perspective, they represent an overall binary size increase of 9\%, 9\%, 12\%, and 20\%, when comparing the size-1 instantiation to the size-256 instantiation of Concat, Remove, Reduce, and Join, respectively.
These increases in binary size are caused by the implicit evidences synthesized by the compiler during implicit resolution.
In our benchmarks, those evidences have no runtime purpose and are simply an artifact of the mechanism in play.
Other applications, such as those supported by the Shapeless library~\citep{sabin2011shapeless}, take advantage of implicit evidence synthesis to provide additional runtime functionalities.

@@

In \Cref{chap:match-types}, we describe the match type reduction algorithm by adding new subtyping rules to \SystemFsub (\Cref{subsec:matches}).
The resulting system is entirely declarative: it doesn't specify \emph{when} to apply those rules.
In this section, we discuss how the timing of match type reduction impacts compilation times.

A correct type system implementation must attempt to reduce match types \emph{as late as possible}, during subtyping.
Indeed, failing to do so could lead to incorrect results in cases where a match type is reducible in the context of the subtyping query but is not reducible in any prior context.

In addition to the reductions attempted during subtyping, the compiler can try to reduce match types earlier in the pipeline, essentially at any point in time.
Doing so is always correct, since the types before and after reduction are mutually subtypes of each other.
Each early attempt of match type reduction is a trade-off: it can either be a small overhead, if the reduction fails, a no-op, if the reduction succeeds but the result is used exactly once, or an optimization, if the reduction succeeds and the result is used more than once.
Note that reducing match types in a lazy fashion would not be a replacement for early reduction attempts, because the subtyping and disjointness checks are context-dependent (\Cref{subsec:caching}).

@@

\noindent
Without any early reductions, this definition will result in a quadratic number of additions. For example, here is the sequence of subtyping queries issued while compiling |GEQ[5,1]|:

@@

Our Scala~3 implementation reduces match types at multiple points during compilation, including right after creating types, during subtyping, and after instantiating parameters of type applications.
This last point is essential to avoid the quadratic behavior of the |GEQ| example.
In \Cref{fig:timingGraphsTooEarlyTooLate}, we show the impact of this optimization on our Reduce benchmark.
In this figure, we compare the compilation times of the match type Reduce benchmark with (S1) and without (S2) the reduction of match types after type parameter instantiation.

@@

In this thesis, we presented three techniques for type-level programming in Scala: implicits, generalized singleton types, and match types.
We showed that programming with implicits can be relatively cumbersome, both for programmers and compilers, which led us to the following question: \emph{can we do better?}
Our main contribution, match types, certainly answers that question positively.
They do provide a simple, first-class solution to type-level programming which is both easy to use and fast to compile.
We formalized match types in a calculus based on \SystemFsub and proved its soundness, which gives us confidence that our approach is correct and sensible.
Our implementation of match types in the Scala~3 compiler makes them accessible to a large audience of programmers and is already in active use.
Beyond the scope of Scala, it will be interesting to see if our design can inspire other general-purpose languages interested in moving their type system towards dependent types.

@@

\input{proofs/inversion-of-subtyping.tex}
\input{proofs/canonical-forms.tex}
\input{proofs/inversion-of-typing.tex}
\input{proofs/minimum-types.tex}
\input{proofs/preservation.tex}
\input{proofs/progress.tex}

>>>>

In this chapter, we evaluate the performance of the various type-level programming techniques presented in prior chapters.
Specifically, we are interested in the scalability of each technique when confronted with large type-level programs, in terms of compilation times (\Cref{sec:compilation-time}) and binary sizes (\Cref{sec:binary-size}).
In \Cref{sec:the-timing-of-match-type-reductions}, we discuss how the timing of match type reduction impacts compilation times.

@@

We evaluate the performance of the three type-level programming techniques presented in this thesis by comparing them on a set of four benchmarks.
Each benchmark consists of a simple type-level operation, which we implemented once with implicits (\Cref{chap:ab-using-implicits}), once with generalized singletons (\Cref{chap:generalizing-singleton-types}), and once with match types (\Cref{chap:match-types}).

@@

  \item \textbf{Remove}\quad
  Removes the last element of a heterogeneous list (inspired by the example presented in \Cref{sec:hlists-remove}).

  \item \textbf{Join}\quad
  Computes the resulting schema after performing a self-join along the last column of a database table (inspired by the example presented in \Cref{sec:motivating-example}).

  \item \textbf{Reduce}\quad
  Computes the dimension of a multidimensional array after dimensional reduction along all axes (inspired by the example presented in \Cref{subsubsec:reduction}).
\end{enumerate}

We obtained the measurements presented in this chapter by averaging independent compilations on warm compilers, which we executed on an i7-7700K processor running Oracle JVM~1.8.0\_212 and Linux.
For the implicit and match type benchmarks, we used the latest version of Scala at the time of writing (3.1.2).
For the generalized singleton benchmarks, we used our prototype, which lives in a branch of the dotty-staging repository\footnote{\texttt{git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7}}.

@@

\Cref{fig:timingGraphsZoomedOut} compares the compilation time of implicits, generalized singletons and match types on our benchmark suite.
The compilation time of implicit-based benchmarks dominates the graphs on each figure, and suggests a quadratic complexity.
These results imply that implicits are ill-suited for large scale type-level computations: spending tens of seconds on a single operation is clearly out of the question for practical applications.

@@

The same measurements are shown in \Cref{fig:timingGraphsZoomedIn} with millisecond y-axes, to focus on the comparison of non-implicit curves.
Each graph follows a similar trend, with singleton-based implementations outperforming their match-type-based counterparts by about a factor of two.
We speculate that this difference is due to the overhead of subtyping, which in inherent to the design of match type.
Our prototype for generalized singletons uses a dedicated type evaluation procedure that is specialized for the task and therefore faster.

We obtained the data points of \Cref{fig:timingGraphsZoomedOut} and~\ref{fig:timingGraphsZoomedIn} using the Java Microbenchmark Harness library with 60 seconds of warmup, 60 seconds of measurement and a single JVM.
As a result, the resulting measurements are precise (99.9\% confidence intervals vary between $\pm$ 0.5 and $\pm$ 2 milliseconds), but subject to variations caused by the unpredictability of the JVM's just-in-time compiler, which explains the bumps visible in \Cref{fig:timingGraphsZoomedIn}.

@@

In \Cref{fig:bytecodeSizeTable}, we show the amount of additional JVM Bytecode generated per benchmark size increment.
The binary sizes of singleton- and match-type-based implementations are stable with respect to the benchmark input size, which is expected given that our benchmarks consist of type-level-only operations.

The binary sizes of implicit-based implementations grow linearly with the input size, by 6 bytes per increment for smaller benchmarks (Concat and Remove), and by 12 and 18 bytes for the larger benchmarks (Reduce and Join, respectively).
To put those numbers in perspective, they represent an overall binary size increase of 9\%, 9\%, 12\%, and 20\%, when comparing the size-1 instantiation to the size-256 instantiation of our Concat, Remove, Reduce, and Join benchmarks, respectively.
These binary size increases are caused by the implicit evidences synthesized by the compiler after implicit resolution.
In our benchmarks, those evidences have no runtime purpose and are simply artifacts of the mechanism in play.
Other applications, such as those supported by the Shapeless library~\citep{sabin2011shapeless}, take advantage of implicit evidence synthesis to provide additional runtime functionalities.

@@

In \Cref{chap:match-types}, we described the match type reduction algorithm by adding new subtyping rules to \SystemFsub (\Cref{subsec:matches}).
The resulting system is entirely declarative: it doesn't specify \emph{when} to apply those rules.
In this section, we discuss how the timing of match type reduction impacts compilation times.

A correct type system implementation must attempt to reduce match types \emph{as late as possible}, that is, during subtyping.
Indeed, failing to do so could lead to incorrect results, for instance, if a match type is reducible in the context of the subtyping query, but is not reducible in any prior context.

In addition to reducing match types during subtyping, the compiler can attempt to reduce match types earlier in the pipeline, essentially at any point in time.
Doing so is always correct: the types before and after reduction are mutually subtypes.
Each early attempt of match type reduction can be viewed as a performance trade-off: it can either introduce a small overhead, if the reduction fails, be a no-op, if the reduction succeeds but the result is used exactly once, or be an optimization, if the reduction succeeds and the result is used more than once.

@@

\noindent
Without any early reductions, this definition results in a quadratic number of additions. For example, here is the sequence of subtyping queries that would be issued while compiling |GEQ[5,1]|:

@@

Our implementation reduces match types at multiple points during compilation, including directly after creating types, during subtyping, and after instantiating parameters of type applications.
This last point is essential to avoid the quadratic behavior of the |GEQ| example.
In \Cref{fig:timingGraphsTooEarlyTooLate}, we show the impact of this optimization on our Reduce benchmark.
In this figure, we compare the compilation times of the match type Reduce benchmark with (S1) and without (S2) the reduction of match types after type parameter instantiation.

@@

In this thesis, we presented three techniques for type-level programming in Scala: implicits, generalized singleton types, and match types.
We showed that programming with implicits can be relatively cumbersome, both for programmers and compilers, which led us to the following question: \emph{can we do better?}
Our main contribution, match types, certainly answers that question positively.
They do provide a simple, first-class solution to type-level programming which is both easy to use and fast to compile.
We formalized match types in a calculus based on \SystemFsub and proved its soundness, which gives us confidence that our approach is correct and sensible.
Our implementation of match types in the Scala~3 compiler is already in active use and makes type-level programming accessible to a large audience of programmers.
Beyond the scope of Scala, it will be interesting to see if our design can inspire other general-purpose languages interested in moving their type system towards dependent types.

@@

\input{proofs/inversion-of-subtyping.tex}
\input{proofs/canonical-forms.tex}
\input{proofs/inversion-of-typing.tex}
\input{proofs/minimum-types.tex}
\input{proofs/progress.tex}
\input{proofs/preservation.tex}
